---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# prinsurf

<!-- badges: start -->
<!-- badges: end -->

The goal of prinsurf is to construct a principal surfaces that are two-dimensional surfaces that pass through the middle of a $p$-dimensional dataset. 

## Installation

You can install the development version of prinsurf from [GitHub](https://github.com/) with:

``` r
library(devtools)
devtools::install_github("RaeesaGaney91/prinsurf")
```

## Example

This is a basic example on a simulated data set:

```{r echo=FALSE}
data2 <- function(n,side)
{
  e1 <- runif(n,0,0.5)
  x <- 2*runif(n) - 1
  theta <- 2*pi*runif(n) - pi
  y <- sin(theta)*sqrt(1-x^3)
  z <- sin(theta)*sqrt(x^2)
  X = matrix(c(x+e1,y+e1,z+e1),nrow=n,ncol=3,byrow=F)
  return(X)
}
X <-  data2(100,1)
library(rgl)
```


```{r example}
library(prinsurf)
surface <- principal.surface(X)
```


```{r echo=FALSE}
rglwidget() # Interactive widget in the RStudio viewer

# Save as a standalone HTML file
writeWebGL(dir = "docs", filename = "docs/surface3d.html")
```

[View Interactive 3D Surface Plot](https://RaeesaGaney91.github.io/prinsurf/surface3d.html)

## Report Bugs and Support
If you encounter any issues or have questions,
please open an issue on the GitHub repository.

