---
title: "Gradient-flow biplot axes"
subtitle: "Theory and implementation"
author: "Raeesa Ganey"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Gradient-flow biplot axes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5.5, fig.align = "center"
)
library(prinsurf)
```

# The axis-reading problem

A principal component biplot (Gower and Hand, 1996) reads variable values
directly: a biplot axis is a straight line calibrated with the variable's
values, and you project each sample orthogonally onto it to recover a
prediction. The geometry works because the axis is embedded in the plane that
contains the fitted values — the PCA plane. The projection is linear and the
arithmetic transparent.

A principal surface is curved. The fitted surface is a smooth 2-dimensional
manifold $\hat S$ embedded in $\mathbb{R}^p$, defined implicitly by the
coordinate functions $\hat f_1, \ldots, \hat f_p$, each estimated by loess on
the surface coordinates $(\lambda_i^{(1)}, \lambda_i^{(2)})$. That is, sample
$i$ maps to the surface point
$$\hat \mathbf{x}_i = (\hat f_1(\lambda_i), \ldots, \hat f_p(\lambda_i)) \in \hat S.$$

Each variable $j$ defines a scalar field over the surface: the function
$\hat f_j : \mathbb{R}^2 \to \mathbb{R}$. Unlike a plane, a curved surface has
no single direction that is "aligned" with the variable; different parts of
the surface may have different gradients. The **gradient-flow axis** for
variable $j$ is the steepest-ascent trajectory of $\hat f_j$, started from the
centre of the sample coordinates. It is smooth (solutions to the gradient-flow
ODE), follows the direction of greatest rate of change, and crosses the
variable's contour lines orthogonally at every point. This is the natural
extension of the biplot axis to a curved surface.

# The gradient field

Given the fitted coordinate functions, we can compute the gradient of $\hat f_j$
at any point $(\lambda_1, \lambda_2) \in \mathbb{R}^2$ by finite differences on
a fine evaluation grid. Let $G_j(\lambda_1, \lambda_2)$ denote the 2-vector
of partial derivatives:
$$G_j = \left( \frac{\partial \hat f_j}{\partial \lambda_1}, \frac{\partial \hat f_j}{\partial \lambda_2} \right).$$

This can be computed on a regular grid (default resolution $N = 60$) by
centred differences on the smoothed field $\hat f_j(\lambda_1, \lambda_2)$.
Points near the boundary of the grid (where one-sided differences are needed)
are handled conservatively: their gradients are set to `NA` and the flow
halts when it reaches one.

# Gradient-flow integration

The gradient-flow trajectory is the solution to the autonomous ODE
$$\frac{d\boldsymbol{\lambda}(t)}{dt} = G_j(\boldsymbol{\lambda}(t)),$$
where $\boldsymbol{\lambda}(t) = (\lambda_1(t), \lambda_2(t))$ is the position
at time $t$. We integrate this numerically using fourth-order Runge-Kutta
(RK4) with step size $h = 0.03$ (default) for up to 800 steps in each
direction (forward and backward) from the starting point.

The starting point is the grid node (after filtering to supported cells) that
is closest to the centroid of the sample coordinates. Integrating backwards as
well as forwards ensures the axis spans the full range where the field is
monotone.

**Why RK4?** The method is robust to the condition of the surface, and because
the loess smooths are locally well-behaved, four stages per step is enough.
Adaptive step-size methods (like `rk45`) would be overkill; fixed small steps
are stable.

The output is a polyline (sequence of vertices) that traces the gradient flow.
Consecutive points that are very close (arc-length difference $< 10^{-6}$) are
filtered as redundant.

# Monotonicity criteria

Not every variable yields a valid axis. Two criteria filter out cases where
monotone reading would be misleading.

## Criterion 1: Interior extremum

If $\hat f_j$ has a local maximum or minimum in the *interior* of the
supported region, strictly bounded by cells with local data support, then no
monotone axis can represent the full field. The contours loop around the
extremum, and any attempt to draw a line through the middle either misses
the extreme value entirely or crosses it non-monotonically.

This is detected by identifying:
- The **boundary** of the supported region: cells with local data support
  that are adjacent to unsupported cells.
- The **interior**: cells with local data support that are not on the boundary.

If any interior cell has a value exceeding the maximum of all boundary cells
by more than $\delta \times (\text{range of } \hat f_j)$ (default $\delta = 0.05$),
or falling below the minimum boundary value by the same margin, the variable
is deferred. The margin $\delta$ prevents false positives at ragged edges
where discrete grid cells create artificial steps.

## Criterion 2: Coverage

Even if monotone, an axis may span too little of the variable's observed
range. We measure this by the **coverage**:
$$\text{coverage} = \frac{\max(\text{axis calibration}) - \min(\text{axis calibration})}
{\max(\hat f_j(\lambda_i)) - \min(\hat f_j(\lambda_i))},$$
i.e., the fraction of the sample range that the axis actually spans. If this
is less than `cover_min` (default 0.55), the axis is deferred because it is
not capturing enough variation to be useful for biplot reading.

Low coverage typically signals a variable with curvature so complex that the
steepest ascent route only touches one flank of the field, leaving the other
flank better visible in contour form.

## When to override

Both criteria are *display* decisions, not verdicts on the variable. A
deferred variable is still fitted by loess, still predictable via
`predict()`, and often the most interesting to understand via its contours.
Pass `defer = FALSE` to `psaxis()` or to `plot()` to force an axis for every
variable with a non-degenerate flow (Criteria 1 and 2 ignored), and inspect
the returned `coverage` to see why deferral was the right call.

# Axis calibration

Once the gradient-flow polyline is in place, each vertex must be labelled with
the variable's value. We evaluate $\hat f_j$ at each vertex:
$$\text{calibration}_k = \hat f_j(\boldsymbol{\lambda}_k),$$
where $\boldsymbol{\lambda}_k$ is the $k$-th vertex of the polyline. These
values are in the working (centred, optionally scaled) units; `plot()` and
`predict()` back-transform them to the original units for display and
prediction.

This design ensures the calibration is consistent with the field the axis
represents: small perturbations of $\boldsymbol{\lambda}$ along the polyline
produce monotone changes in the calibration, because the polyline follows the
gradient direction.

# Reading the axis

To predict variable $j$ for a sample at $\boldsymbol{\lambda}_i$, orthogonally
project $\boldsymbol{\lambda}_i$ onto the axis polyline. This gives the closest
point on the axis (in Euclidean distance), and the calibration value at that
point is the prediction. The orthogonal projection method is:

For each segment of the polyline (from vertex $k$ to $k+1$),
parametrize it as $\mathbf{a} + t (\mathbf{b} - \mathbf{a})$ for $t \in [0, 1]$.
The foot of the perpendicular from $\boldsymbol{\lambda}_i$ to this segment is
at
$$t^* = \frac{(\boldsymbol{\lambda}_i - \mathbf{a}) \cdot (\mathbf{b} - \mathbf{a})}{\lVert \mathbf{b} - \mathbf{a} \rVert^2},$$
clamped to $[0, 1]$. Compute this for every segment, find the segment with
minimum distance, and linearly interpolate the calibration value.

**Why orthogonal projection?** It is symmetric in how it treats a sample
relative to the axis, robust to axis curvature, and identical to the reading
rule of a straight-line biplot axis when the axis is indeed straight. It is
also the definition used in Gower's predictive-biplot tradition for
computing axis predictive error.

# Comparison to alternatives

Two other designs were considered and rejected.

**Back-projection to the principal-component plane.** Take the initial PCA
plane, project each sample onto the nearest surface point, then project *that*
back onto the plane. This loses information: the back-projected point is no
longer on the surface, and the reading is not about the surface field at all.

**Independent calibration by ranking.** Find the sample with the highest
variable value, the second-highest, etc., and mark those positions on the axis
as calibration points. This is brittle when samples cluster, prone to
outlier-driven calibration, and loses the connection between the axis's
geometric path and the field it represents. Most critically, the axis itself
is determined by gradient flow (a property of the fitted surface), while the
calibration is data-driven; decoupling them invites inconsistency.

The gradient-flow approach integrates the geometry and the data naturally: the
axis follows the field, and the calibration represents that same field.

# Implementation details

**Grid resolution** (`N`, default 60): Higher values give smoother gradients
but more computation. 60 is a good balance; go to 80–100 if you have many
variables and a fast machine, or 40–50 for exploratory work.

**Step size** (`h`, default 0.03): Smaller values trace the flow more finely
but integrate longer. 0.03 is conservative; if your axes look jagged, decrease
it slightly. If they look over-smooth or integration takes too long, increase
it.

**Max steps** (default 800): In each direction from the start. The flow stops
early if it leaves the supported region or hits a degenerate point. 800 is
plenty for practical datasets; lower it only if you are impatient.

**Boundary margin** (`delta`, default 0.05): If interior extrema exceed
boundary values by less than this fraction of the range, they are tolerated.
Increase to be stricter (defer more variables); decrease to be more lenient.

**Coverage threshold** (`cover_min`, default 0.55): Variables with axis
coverage below this are deferred. Lower values accept more variables with
narrow coverage; higher values defer conservatively.

# Example: the blanket surface

We use the blanket data to visualize the gradient-flow principle.

```{r blanket-setup}
set.seed(1)
n  <- 200
s  <- runif(n, -1, 1)
t  <- runif(n, -1, 1)
e  <- function() runif(n, -0.05, 0.05)
x1 <- s + e()
x2 <- t + e()
x3 <- 0.7 * s + 0.9 * t^2 + e()
X  <- cbind(x1, x2, x3)
colnames(X) <- c("x", "y", "z")
fit_b <- prinsurf(X, verbose = FALSE)
```

The $z$ coordinate has parabolic contours in the latent $(s, t)$ space.
Its gradient-flow axis should follow the ridge of the parabola, bending
smoothly. Let's inspect it:

```{r blanket-z-axis}
ax <- psaxis(fit_b, "z")
print(ax)
```

The axis has `r nrow(ax$axis)` vertices and spans calibration range
`[`r round(min(ax$cal), 2)`, `r round(max(ax$cal), 2)`]` in working units,
with `r sprintf("%.2f", ax$coverage)` coverage. Plot it on the surface:

```{r blanket-z-plot}
plot(fit_b, vars = "z", contours = "z", col_contour = "lightblue")
```

The axis (black, with tick marks) follows the valley of the contours,
staying orthogonal to them. Its path is smooth and monotone in the displayed
calibration, even though the contours themselves are curved. This is exactly
what the gradient flow achieves: it finds the direction of steepest ascent
and follows it faithfully.

Compare this to what a forced straight axis would show (by setting
`defer = FALSE`):

```{r blanket-z-straight, eval=FALSE}
ax_straight <- psaxis(fit_b, "z", defer = FALSE)
# This would show an axis that tries to be linear but fails to track
# the parabolic field — gaps in coverage, or overshooting into contours.
```

The gradient-flow axis is better because it adapts to the geometry.

# References

Gower, J.C. and Hand, D.J. (1996). *Biplots*. Chapman & Hall, London.

Gower, J.C., Lubbe, S. and le Roux, N. (2011). *Understanding Biplots*.
Wiley, Chichester.

Hastie, T. and Stuetzle, W. (1989). Principal curves.
*Journal of the American Statistical Association*, **84**, 502–516.

Le Roux, B. and Rouanet, H. (2004). *Geometric Data Analysis: From
Correspondence Analysis to Structured Data Analysis*. Kluwer Academic
Publishers.
