---
title: "Principal-surface biplots with gradient-flow axes"
author: "Raeesa Ganey"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Principal-surface biplots with gradient-flow axes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6.5, fig.height = 5.2, fig.align = "center")
set.seed(1)
```

## What this package does

A biplot equips a low-dimensional scatter of samples with calibrated *axes*, so
that a variable's value for a sample can be recovered by projecting the sample
onto the axis and reading the scale. For a **principal surface** -- a smooth
curved two-dimensional manifold fitted through the data (Hastie & Stuetzle,
1989) -- the classical axis is the *back-projection* axis of nonlinear biplots:
for each marker value one point of that value's contour is chosen,
independently of the others, and the points are joined in order. The result is
continuous but kinked.

Those kinks are not merely unsightly. An axis that doubles back on itself makes
orthogonal projection *ill-defined* -- a sample can have two feet on the axis
with different readings -- and prediction can be worse than simply using the
variable's mean. (The PhD thesis underlying this package documents exactly such
a case.)

`prinsurf` replaces that axis with the **gradient-flow axis**. For variable
\(j\) with fitted coordinate function \(\hat f_j\), the calibrated axis is the
steepest-ascent trajectory of \(\hat f_j\),
\[
  \frac{d\gamma_j}{ds}=\frac{\nabla \hat f_j(\gamma_j(s))}{\lVert\nabla \hat f_j(\gamma_j(s))\rVert},
  \qquad \gamma_j(0)=\bar{\boldsymbol\lambda},
\]
started at the centroid of the sample coordinates. Three properties are proved
in the accompanying theory note: the axis is **smooth** (Picard--Lindelöf), it
crosses the variable's contour lines **orthogonally** with monotone calibration,
and it **reduces exactly to the linear PCA-biplot axis when the surface is
flat**. The last property is what gives it standing: this is not a separate
device bolted onto principal surfaces, but the canonical generalisation of the
prediction axis to curved surfaces.

## Fitting a surface and drawing the biplot

```{r fit-iris}
library(prinsurf)

fit <- prinsurf(iris[, 1:4], scale = TRUE)   # standardise heterogeneous units
fit
```

`plot()` draws the sample coordinates with a calibrated gradient-flow axis for
each variable. Setting `defer = FALSE` draws an axis for *every* variable (we
return to deferral below).

```{r plot-iris}
plot(fit, group = iris$Species, defer = FALSE)
```

Each axis is a smooth curve, calibrated in the variable's own units, with ticks
at convenient values and the variable name at the high end. A single variable's
axis is available directly:

```{r psaxis}
ax <- psaxis(fit, "Petal.Length")
ax
kink_max(ax$axis)        # largest turning angle, in degrees: a smooth axis is small
```

`ax$axis` is the polyline (many closely spaced vertices), `ax$cal` the
calibration at each vertex, and `ax$g1`, `ax$g2`, `ax$M` the grid of fitted
values used for contours.

## When a single axis is not enough

A single calibrated axis only makes sense when \(\hat f_j\) is monotone over the
surface. If the variable folds -- an interior peak or basin, i.e. closed
contours -- no single axis can carry it: every axis would assign one value to a
whole closed loop. `prinsurf` detects this with two criteria:

* **Boundary-extremum** (Criterion 1): an interior value beyond the entire
  support-boundary range certifies a genuine interior peak or basin.
* **Coverage** (Criterion 2): after the axis is built, it must span at least
  `cover_min` (default 0.55) of the variable's range over the samples;
  otherwise the flow has stalled at a local extremum and the axis covers only
  part of the variable.

A variable failing either is *deferred* and read from its contour lines instead.
The half-sphere below is the canonical illustration: its height coordinate has
concentric circular contours (an interior maximum), while the two horizontal
coordinates are monotone.

```{r halfsphere}
## upper cap of a sphere: x = height (folded), y, z = horizontals (monotone)
n <- 200; u <- 2 * runif(n) - 1; th <- 2 * pi * runif(n) - pi
S <- cbind(x = u, y = sin(th) * sqrt(1 - u^2), z = cos(th) * sqrt(1 - u^2))
S <- S[S[, "x"] > -0.4, ]                 # keep the cap
S <- sweep(S, 2, colMeans(S))             # centre
sph <- prinsurf(S, max.iter = 8)

vapply(colnames(S), function(v) isTRUE(psaxis(sph, v)$monotone), logical(1))
```

`x` is deferred; `y` and `z` get axes. We draw the monotone axes and overlay the
contours of the deferred `x`:

```{r halfsphere-plot}
plot(sph, contours = "x")
```

The closed contours of `x` are exactly the structure that defeats a single axis.
The biplot is therefore *mixed by construction*: monotone variables carry
calibrated axes read by projection, folded variables are read from contours --
mirroring the classical split between axis prediction and contour prediction,
but now decided automatically and for a stated reason.

## Reading and prediction: an honest account

It is worth being precise about two different notions of "prediction", because
they behave differently.

**Sample predictivity** asks how well the surface reconstructs each *whole*
sample, \(\hat{\boldsymbol x}_i = \hat f(\boldsymbol\lambda_i)\); it is read from
the sample's *position* and does not use any single axis. Here the curved
surface beats a PCA biplot, because it captures nonlinear structure the flat
plane cannot.

```{r sample-pred}
ps  <- mean(predictivity(fit))

## PCA biplot (rank-2) sample predictivity on the same standardised data
Z   <- scale(as.matrix(iris[, 1:4]))
V2  <- svd(Z)$v[, 1:2]; Zhat <- Z %*% V2 %*% t(V2)
pca <- mean(1 - rowSums((Z - Zhat)^2) / rowSums(Z^2))

c(principal_surface = round(ps, 3), pca_biplot = round(pca, 3))
```

**Axis reading** asks how well a *single* axis recovers a variable by orthogonal
projection. On a curved surface this is a one-dimensional summary, and it is
lossy wherever the contours bend: the perpendicular foot of a sample lands at a
fitted value away from the sample's own. The per-axis root-mean-square reading
error quantifies this.

```{r axis-error}
axis_predictive_error(fit)              # PS: RMS reading error per axis (NA = deferred)

## PCA biplot per-variable reading error (rank-2), same standardised units
sqrt(colMeans((Z - Zhat)^2))
```

A PCA biplot's straight axis, by contrast, reads the *full* rank-2 value
(projection onto a straight loading returns both coordinates' contribution), so
on a near-linear data set such as iris it reads individual variables more
accurately than a curved one-dimensional axis does. This is geometry, not a
defect: exact projection reading requires straight contours.

The practical reading of all this: the gradient-flow axis is a trustworthy
**display** of a variable's direction and approximate scale -- well-defined,
smooth, orthogonally calibrated -- and a principled **diagnostic** for which
variables a single axis can carry. Precise values for folded variables come from
contours. The surface's *predictive* advantage over PCA lives in the
position/contour reading, not in projection onto one curved axis.

## The linear limit

When the data are essentially two-dimensional the surface is flat, and the
gradient-flow axis collapses onto the straight PCA-biplot axis. We can see this
directly: generate near-planar data, and the axis is almost straight (tiny
kink) and points along the variable's loading direction.

```{r linear-limit}
m <- 150; a <- runif(m, -1, 1); b <- runif(m, -1, 1)
L <- cbind(v1 = a, v2 = b, v3 = 0.6 * a - 0.4 * b) + matrix(rnorm(3 * m, 0, 0.01), m, 3)
flat <- prinsurf(L, max.iter = 6)
kink_max(psaxis(flat, "v1")$axis)     # near 0 degrees: the axis is straight
psaxis(flat, "v1")$coverage           # near 1: it spans the variable
```

## Where this is going

The surface represents the data better than PCA on sample predictivity, yet a
single one-dimensional axis read by orthogonal projection cannot realise that
advantage for curved variables -- the same limitation faced by every nonlinear
biplot. The natural next step is a *reading rule* that uses a sample's full
position (not only its perpendicular foot on the axis) so that a single readable
per-variable object predicts at the surface's accuracy. Such an element -- an
axis that is read like an axis but predicts like the surface -- would turn the
present contribution (a well-defined, calibrated, diagnostic-backed axis) into
one that also predicts what a PCA biplot cannot. That is the open direction this
package is meant to support exploring.

## Reference

Theory, propositions and proofs are in the accompanying note `theory.pdf`;
the underlying methodology is the principal surface of Hastie & Stuetzle (1989)
and the nonlinear/predictive biplot framework of Gower, Gardner-Lubbe & le Roux
(2011), *Understanding Biplots*.
